---
title: "Univariate Descriptive"
author: "Andrea Mascaretti"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```

# Curse of dimensionality
## Introduction
Given $N$ random independend points with uniform distribution in the ball $\mathcal{B}_{p}\left[0,\,1\right] \in \mathbb{R}^{p}$, compute the density of the distance, the expected value and the median of the point closest to the origin.
The higher $p$, the bigger the values of the support of the aforementioned distribution





## Density
Density of the distribution of the distance of the $N$ closest points $x_i$ such that $x_i \sim \mathcal{U}\left(\mathcal{B}_{p}\left[0,\,1\right]\right)$.

```{r density-function}
dmin <- function(x, p, N) {
  p * N * (1 - x^p)^(N - 1) * x^(p - 1)
}
```




### Median
We create a function for the median of the distribution of the distance of the $N$ closest points $x_i$ such that $x_i \sim \mathcal{U}\left(\mathcal{B}_{p}\left[0,\,1\right]\right)$.

```{r density-function-median}
medmin <- function(p, N) {
  (1 - .5^(1 / N))^(1 / p)
}
```



### Mean
We set the mean of the distribution of the distance of the $N$ closest points $x_i$ such that $x_i \sim \mathcal{U}\left(\mathcal{B}_{p}\left[0,\,1\right]\right)$.

```{r}
meanmin <- function(p, N) {
  sdmin <- function(x) {
    x * dmin(x, p, N)
  }
  integrate(sdmin, 0, 1)
}
```




## Visualising the curse of dimensionality
### Density
We plot the density for varying $p$ (the space dimension) and set $N = 100$.

```{r plots}
x <- seq(0, 1, .01)


plot(x, dmin(x, 1, 100),
  type = "l", main = "Density of MinDistance",
  xlab = "x", ylab = "density", ylim = c(0, 20), lwd = 3
)
lines(x, dmin(x, 5, 100), col = "green", lwd = 3)
lines(x, dmin(x, 10, 100), col = "red", lwd = 3)
lines(x, dmin(x, 20, 100), col = "blue", lwd = 3)
legend(
  x = .8, y = 20, legend = c("p=1", "p=5", "p=10", "p=20"),
  fill = c("black", "green", "red", "blue")
)
```





### Median

We plot the median for varying $p$ (the space dimension) and set $N = 100$.

```{r}
plot(1:20, medmin(1:20, 100),
  main = "Median of MinDistance (N=100)",
  xlab = "Dimension p", ylab = "Median(MinDistance)", ylim = c(0, 1)
)
```


```{r}
mediana <- c()
for (p in 1:5) mediana <- c(mediana, medmin(p, 100^p))
p <- 1:5


plot(p, mediana,
  main = "Median of MinDistance (N=100^p)",
  xlab = "Dimension p", ylab = "Median(MinDistance)", ylim = c(0, 1)
)

```




### Mean


We plot the mean for varying $p$ (the space dimension) and set $N = 100$.

```{r}
media <- c()
for (p in 1:20) media <- c(media, meanmin(p, 100)$value)
p <- 1:20

plot(p, media,
  main = "Expected value of MinDistance (N=100)",
  xlab = "Dimension p", ylab = "E(MinDistance)", ylim = c(0, 1)
)

```
