---
title: "Classification - Human Data"
author: "Andrea Mascaretti"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(knitr)
library(learnr)
knitr::opts_chunk$set(echo = TRUE, exercise = FALSE)
```

# Classification
## Preliminary Analysis
We are going to attempt to classify the gender of a group of humans, based on their heights.
```{r load-libary}
library(MASS)
library(class)
library(kknn)
library(rpart)
```

```{r read-data}
mydata <- read.table("../data/cleaneddata.txt", header = T, sep = ",")
mydata.gender <- mydata[, 1]
mydata.biosocial <- mydata[, 3:8]

group <- mydata.gender
feature1 <- mydata.biosocial[, "height"]
```

We also make a plot.
```{r plot-data}
plot(feature1, col = group)
```



## LDA and QDA
We fit the models.
```{r lda-qda-fit}
lda.fit <- lda(group ~ feature1)
qda.fit <- qda(group ~ feature1)
```

Estimate the posterior group probabilities provided by the two models for a subject whose height is 175 cm.
```{r fit-175}
predict(lda.fit, data.frame(feature1 = 175))
predict(qda.fit, data.frame(feature1 = 175))
```


## Grid
We estimate posterior probabilities on a regular grid provided by the two models
```{r fit-grid}
x <- data.frame(feature1 = seq(min(feature1), max(feature1), length = 100))
lda.posterior <- predict(lda.fit, x)$posterior
qda.posterior <- predict(qda.fit, x)$posterior
```


### Plot posterior
We plot the probabilities we obtain using LDA.
```{r}
# LDA Group 1
plot(x[, 1], lda.posterior[, 1], type = "l", col = "blue", xlab = "x", ylab = "estimated posterior")
# LDA Group 2
points(x[, 1], lda.posterior[, 2], type = "l", col = "red")
```

and overplot those obtained with QDA.
```{r}
# LDA Group 1
plot(x[, 1], lda.posterior[, 1], type = "l", col = "blue", xlab = "x", ylab = "estimated posterior")
# LDA Group 2
points(x[, 1], lda.posterior[, 2], type = "l", col = "red")
# QDA Group 1
points(x[, 1], qda.posterior[, 1], type = "l", col = "blue", lty = 2, xlab = "x", ylab = "estimated posterior")
# QDA Group 2
points(x[, 1], qda.posterior[, 2], type = "l", col = "red", lty = 2)
```



## k-Nearest Neighbour Classifier
We fit the model on a regular grid.
```{r knn-fit}
x <- data.frame(feature1 = seq(min(feature1), max(feature1), length = 100))
(knn.fitted <- knn(train = feature1, test = x, cl = group, k = 30, prob = T))
```

We need to chose $k$.
```{r}
train.cv <- train.kknn(group ~ feature1,
  data = data.frame(feature1, group),
  kmax = 40, scale = F, kernel = "rectangular"
)
plot(train.cv)
```




## CART
We fit the model
```{r cart-fit}
rpart.fit <- rpart(group ~ feature1)
```


and stimate the posterior group probabilities for a subject whose height is 175cm.
```{r cart-175}
predict(rpart.fit, data.frame(feature1 = 175))
```

We compute the values on a grid.
```{r cart-grid}
x <- data.frame(feature1 = seq(min(feature1), max(feature1), length = 100))
data.frame(x = x, female = predict(rpart.fit, x)[, 1], male = 1 - predict(rpart.fit, x)[, 1])
```



## Multivariate Analysis
We now consider two classes: `Female` and `Male` and six features.
```{r social-data}
group <- mydata.gender
features <- mydata.biosocial
pairs(features, col = group)
```



We fit the QDA.
```{r social-qda}
qda.fit <- qda(group ~ ., features)
```


Estimate the posterior group probabilities for the following subject:
```{r subject-qda}
kable(data.frame(age = 23, height = 175, weight = 70, phone = 100, facebook = 1000, instagram = 500))
```
```{r social-qda-subject}
predict(qda.fit, data.frame(age = 23, height = 175, weight = 70, phone = 100, facebook = 1000, instagram = 500))
```


We also estimate the posterior probabilities for the subjects in the original dataset.
```{r social-qda-posterior}
data.frame(predict(qda.fit))
```



We compute the confusion matrix.
```{r social-qda-confusion}
kable(table(True.class = group, Predicted.class = predict(qda.fit)$class))
```

and the *apparent* Accuracy and Error Rate.
```{r social-qda-accuracy}
Conf <- table(True.class = group, Predicted.class = predict(qda.fit)$class)
Accuracy <- sum(diag(Conf)) / sum(Conf)
ErrorRate <- 1 - Accuracy
```




## LOO CV
We compute the confusion matrix
```{r social-qda-confusion-loo}
kable(table(True.class = group, Predicted.class = predict(qda.fit, method = "looCV")$class))
```

and the *apparent* Accuracy and Error Rate.
```{r social-qda-accuracy-error-rate-loo}
ConfCV <- table(True.class = group, Predicted.class = predict(qda.fit, method = "looCV")$class)
kable(ConfCV)
(AccuracyCV <- sum(diag(ConfCV)) / sum(ConfCV))
(ErrorRateCV <- 1 - AccuracyCV)
```




## kNN
We fit the model on the original dataset, after rescaling.
```{r rescale}
rescale <- function(x) {
  (x - min(x)) / (max(x) - min(x))
}
```
```{r rescale-data}
mydata.biosocial.rescaled <- as.data.frame(lapply(mydata.biosocial, rescale))

group <- mydata.gender
features <- mydata.biosocial.rescaled
```


We fit the algorithm.
```{r social-knn}
knn.fitted <- knn(train = features, test = features, cl = group, k = 11, prob = T)
```

Estimate the posterior group probabilities for the following subject:
```{r subject-knn}
kable(data.frame(age = 23, height = 175, weight = 70, phone = 100, facebook = 1000, instagram = 500))
```
```{r knn-subject}
(knn.fitted <- knn(train = features, test = data.frame(age = 23, height = 175, weight = 70, phone = 100, facebook = 1000, instagram = 500), cl = group, k = 11, prob = T))
```


We also select the value of $k$ using leave-one-out cross-validation.
```{r knn-loo}
train.cv <- train.kknn(group ~ .,
  data = data.frame(features, group),
  kmax = 50, scale = F, kernel = "rectangular"
)
plot(train.cv)
```



We plot the confusion matrix
```{r knn-confusion}
knn.fitted <- knn(train = features, test = features, cl = group, k = train.cv$best.parameters$k, prob = T)
kable(table(True.class = group, Predicted.class = knn.fitted))
```


We also compute the *apparent* Accuracy and Error Rate.
```{r knn-accuracy-error-rate}
Conf <- table(True.class = group, Predicted.class = knn.fitted)
kable(Conf)
(Accuracy <- sum(diag(Conf)) / sum(Conf))
(ErrorRate <- 1 - Accuracy)
```






## CART
WE fit the model and plot the results.
```{r cart-fit-multivariate}
group <- mydata.gender
features <- mydata.biosocial
rpart.fit <- rpart(group ~ ., data = features)
```
```{r cart-plot}
plot(rpart.fit)
text(rpart.fit)
```



Estimate the posterior group probabilities for the following subject:
```{r subject-cart}
kable(data.frame(age = 23, height = 175, weight = 70, phone = 100, facebook = 1000, instagram = 500))
```
```{r cart-subject-fit}
predict(rpart.fit, data.frame(age = 23, height = 175, weight = 70, phone = 100, facebook = 1000, instagram = 500))
```

We estimate the probabilities on the original dataset.
```{r cart-dataset}
data.frame(predict(rpart.fit))
```


We also compute the confusion matrix, the *apparent* Accuracy and the error rate.
```{r random}
kable(table(True.class = group, Predicted.class = predict(rpart.fit, type = "class")))

# Compute (apparent) Accuracy and Error Rate
Conf <- table(True.class = group, Predicted.class = predict(rpart.fit, type = "class"))
kable(Conf)
(Accuracy <- sum(diag(Conf)) / sum(Conf))
(ErrorRate <- 1 - Accuracy)
```