---
title: "Univariate Descriptive"
author: "Andrea Mascaretti"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```



# Multivariate Statistics


## Introduction
We begin this laboratory by pointing to some useful resources regarding multivariate statistics:

   * a dedicate page on the [R project website](https://cran.r-project.org/web/views/Multivariate.html)
   * the [Little Book of R for Multivariate Analysis](https://little-book-of-r-for-multivariate-analysis.readthedocs.io/en/latest/)


Before beginning the analysis, we load the required libraries.
```{r load-libraries}
library(plotly)
library(corrplot)
library(RColorBrewer)
library(gclus)
library(hexbin)
library(scatterplot3d)
```

## MTCARS data
The data was extracted from the 1974 Motor Trend US magazine. It comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973-74 models).



### Loading data
```{r mtcars-load}
data(mtcars)
mtcars
attach(mtcars)
```





### Pairs
We plot the data.
```{r mtcars-plot}
plot(mtcars)
```

There too many variables! We try to select the interesting ones.
```{r mtcars-plot-reduced}
plot(mtcars[1:7])
```

Can you see any kind of dependence? Among which variables?





### 3d scatter plot
We also plot a 3d scatter plot.
```{r mtcars-plot-3d}
p <- plot_ly(mtcars, x = ~mpg, y = ~wt, z = ~qsec) %>%
  add_markers() %>%
  layout(scene = list(xaxis = list(title = 'mpg'),
                     yaxis = list(title = 'wt'),
                     zaxis = list(title = 'qsec')))

p
```






### Star plots
We make a star plots as well.
```{r mtcars-stars-first}
stars(mtcars[, 1:7],
  key.loc = c(14, 2),
  main = "Motor Trend Cars : stars(*, full = F)", full = FALSE
)
```


```{r mtcars-stars-second}
stars(mtcars[, 1:7],
  key.loc = c(14, 1.5),
  main = "Motor Trend Cars : full stars()", flip.labels = FALSE
)
```




### Location and dispersion indices

We have the mean vector,
```{r mtcars-mean}
colMeans(mtcars)
apply(mtcars,2,mean)
```



We also compute the Variance-Covariance Matrix,
```{r mtcars-var}
var(mtcars)
```


the correlation Matrix (with each entry within $[-1, 1]$),
```{r mtcars-corr}
cor(mtcars)
```

and we round it to make it easier to read it.
```{r mtcars-corr-rounded}
round(cor(mtcars), 2) 
```






### Scatterplot highlighting correlations
We now make a corrplot: positive correlations are displayed in blue and negative correlations in red.
Colour intensity and the size of the circle are proportional to the correlation coefficients.
In the right side of the correlogram, the legend color shows the correlation coefficients and the corresponding colors.

```{r mtcars-corrplot}
M <- cor(mtcars)
corrplot(M, type = "upper", order = "hclust", col = brewer.pal(n = 8, name = "RdYlBu"))
```




Notice that we can customise the plot.
```{r mtcars-corrplot-custom}
corrplot(M, method = "circle", type = "upper")
corrplot(M, method = "pie", type = "upper")
corrplot(M, method = "ellipse", type = "upper") # nice
corrplot(M, method = "color", type = "upper")
corrplot(M, method = "number", type = "upper")
```







### Heatmap
We now visualise using a heatmap and the display of a clustering tree.
```{r mtcars-heatmap}
col <- colorRampPalette(c("blue", "white", "red"))(20)
heatmap(x = M, col = col, symm = TRUE)
```




## Reduce data set
We now consider only a few variables.

```{r df}
(data <- mtcars[, c(1, 3, 5, 6)])
```

We find the mean vector,
```{r df-mean}
colMeans(data)
```


the variance matrix
```{r df-var}
var(data)
```


```{r df-covar}
cor(data)
```





### Basic Scatterplot Matrix
We make a basic scatterplot matrix.
```{r df-scatterplot}
pairs(data, main = "Simple Scatterplot Matrix")
```



```{r df-scatterplot-continued}
data.r <- abs(cor(data)) # get absolute values of correlations
data.col <- dmat.color(data.r) # get colors
cpairs(data, panel.colors = data.col, gap = .5, main = "Variables Ordered and Colored by Correlation")
```



### 3d Plots


We make some 3d plots!
One is interactive:
```{r}
p <- plot_ly(mtcars, x = ~mpg, y = ~wt, z = ~qsec) %>%
  add_markers() %>%
  layout(scene = list(xaxis = list(title = 'mpg'),
                     yaxis = list(title = 'wt'),
                     zaxis = list(title = 'qsec')))

p
```


And here we project on a plane
```{r}
scatterplot3d(wt, disp, mpg, main = "3D Scatterplot")
```


and add some colour.
```{r}
scatterplot3d(wt, disp, mpg, pch = 16, highlight.3d = TRUE, type = "h", main = "3D Scatterplot")
```






## High Density Scatterplot with Binning
We now simulate some data.

```{r simulate-data}
x <- rnorm(1000)
y <- rnorm(1000)
```


We make some bins: we count the points falling in each occupied cell.
```{r simulate-bin}
bin <- hexbin(x, y, xbins = 50)
summary(bin)
```


And plot it!
```{r simulate-bin-plot}
plot(bin, main = "Hexagonal Binning")
```




## Back to MTCARS data
We plot the data using the bins.
```{r plot-scatter}
plot(data$mpg, data$wt, xlab = "mpg", ylab = "wt", main = "Scatterplot")
```


```{r plot-bin}
bin <- hexbin(data$mpg, data$wt, xbins = 10, xlab = "mpg", ylab = "wt")
plot(bin)
```






## Exercise
We are going to use the `wine.txt` file.

We first collect the data.
```{r wine-data}
wine <- read.table("../data/wine.txt")
wine <- wine[, -1]
colnames(wine) <- c(
  "ID", "Alcohol", "Malic acid", "Ash", "Alcalinity of ash", "Magnesium",
  "Total phenols", "Flavanoids", "Nonflavanoid phenols", "Proanthocyanins",
  "Color intensity", "Hue", "OD280/OD315", "Proline"
)
wine
```


Describe the data and give suitable visual representation of the variables contained in the dataset. Select the relevant variables.
```{r wine-data-description, exercise = TRUE}

```
```{r wine-data-description-hint-1}
pairs(wine[2:14]) # Is this useful?
```
```{r wine-data-description-hint-2}
pairs(wine[2:6])
```


Compute the main location and dispersion indices.
```{r wine-data-indices, exercise = TRUE}

```
```{r wine-data-indices-hint-1}
apply(wine[2:14],2,mean)  
```
```{r wine-data-indices-hint-2}
var(wine)
```
```{r wine-data-indices-hint-3}
cor(wine)
```



Select a couple of continuous and/or categorical variables. Analyse them separaetely, as in the univariate case we have gone through in the previous lessons.
```{r wine-data-univariate, exercise = TRUE}

```