---
title: "Graphics in Base R"
author: "Andrea Mascaretti"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```


# Graphics in R

## Introduction

In this tutorial, we are going to focus on plots and graphics. We are going to follow the [official R documentation](https://cran.r-project.org/)

Graphical facilities are an important part of the R environment. They also play a crucial role in the vast majority of statistical enquiries. 

At start up, R initiates a *device driver*, which opens a *graphics window* for the display of graphics.

There are three main groups of R plotting commands:
  - High Level
    Plotting functions that create new plots on the graphic device, usually including axes, labels, titles and so on.
  - Low Level
    Plotting functions that add more information to an existing plot, such as extra points or lines (or labels, etc...).
  - Interactive
    Functions that allow you to interact with a graph to extract information or to add it.

We are here going to focus only on the base graphics. However, there are some alternatives. For instance, a separate sub-system is contained in the **grid** packages: it is both more powerful and complicated. Another option is the [**lattice**](http://lattice.r-forge.r-project.org/): it builds on **grid** and expands its capacities.
An interesting alternative is the [ggplot2](https://ggplot2.tidyverse.org/) package: it contains a set of function to create rather beautiful plots.


## High-Level

### The `plot` function

One of the function that is used most often is the `plot()` function. It is a *generic* function: the type of plot produced depends on the type or *class* of the first argument.

   * `plot(x, y)` or `plot(xy)`
     - If `x` and `y` are vectors, the command produces a scatterplot of `y` against `x`. The same effect is obtained by providing a list containing to elements `x` and `y` or a two-column matrix.
     
   * `plot(x)`
      - If `x` is a time series, then a time-series plot is created. If `x` is a vector, then a plot containing the values in the vector against its index is produced. If `x` is a complex vector, then a plot of imaginary versus real part is created.
      
   * `plot(f)` or `plot(f, y)`
     - If `f` is a is a factor object, `y` is a numeric vector. The first one creates a bar plot of `f`, whereas the second produces boxplots of `y` for each level of `f`.
     
   * `plot(df)` or `plot(~ expr)` or `plot(y ~ expr)`
     - In this case, `df` is a data frame, `y` is an object of any kind and `expr` is a list of objects, separated by `+`. The first two forms produce distributional plots of the variables in a data frame. (first form) or of a number of named objects (second form). The third firm plots `y` against every object named in `expr`.


**Exercise:** Experiment a bit with the iris dataset!
```{r high-level, exercise = TRUE}
data("iris")

```


### Displaying multivariate data

R provides two very functions to represent multivariate data: `pairs()` and `coplot()`.

Let us assume that we wish to plot `X`, a data frame (or numeric matrix).

The command `pairs(X)` produces a pairwise scatterplot matrix of the variables defined by the columns of `X`, thus resulting in $n(n-1)$ plots. The plots are arranged in a matrix with the rows and columns  that are scale constant.

However, when three or four variables are involved, the `coplot()` function may be more useful.
Let us assume that we have `a` and `b`, both numeric vectors, and `c`, which is either a numeric vector or a factor object. The command `coplot(a ~ b | c)` produces scatterplots of `a` against `b`, for given values of `c`. Whenever `c` is a factor, this means that `a` is plotted against `b` for the different values of `c`. If `c` is numeric, then it is divided into a number of *conditioning intervals* and, for each interval, `a` is plotted against `b` for the values of `c` within the interval.
It is possible to also write `coplot(a ~ b | c + d)`, to produce scatterplots of `a` against `b` for every joint conditioning interval of `c` and `d`.
Notice that even though both functions produce scatterplots by default, it is possible to obtain different plots by setting the value of the `panel =` argument of the functions.
An interesting results is obtained setting `panel = panel.smooth()` in the `pairs` function.


**Exercise:** Try the various functions on the Iris dataset.
```{r multivariate-data, exercise = TRUE}
data(iris)

```


### Display Graphics

Other high-level graphics functions that are worth mentioning are `qqnorm()`, `qqline()`, `qqplot()`, `hist()`, `dotchart()`

### Arguments to high-level functions

One of the most important is `type =`. It takes different values:

   * `type = "p"`
      - Plot individual points
    
   * `type = "l"`
      - Plot lines
    
   * `type = "b"`
      - Plot points connected by lines (i.e. *both*)
    
   * `type = "o"`
      - Plot points overlaid by lines
    
   * `type = "h"`
      - Plot vertical lines from points to the zero axis: *high-density*
    
   * `type = "s"`, `type = "S"`
      - Step-function plots. In the first form, the top of the vertical defines the point; in the second the bottom.
    
   * `type = "n"`
      - No plotting at all. It may be useful for successive iterations.

Labels are set by using the `xlab = ` and `ylab = ` arguments. A string is expected afterwards. In a similar fashion, the title and the subtitles are set with `main = ` and `sub = `


**Exercise:** Plot different line types and points.
```{r plot-lines, exercise = TRUE}

```


## Low-level

Sometimes high-level functions are simply not enough. Low-level plotting commands were created just for that: they can be used to add extra information to the current plot.

For instance, `points(x, y)` adds points to a graph, `lines(x, y)` adds points or connected lines (depending on the `type = ` argument), `abline(a, b)` adds line of slope `b` and intercept `a` and so on.


*Exercise*: Get creative! Create a plot of the iris dataset and add some points and lines.
```{r low-level, exercise = TRUE}
data(iris)
```


## Interactive
We are not going into the details of this particular set of functions. We only cover the `locator(n, type)` function. This function waits for the user to select locations on the current plot using the left mouse button. It continues until `n` points have been selected (default is 512) or another mouse button is pressed. The `type` allows for plotting at the selected point.
An example may be `text(locator(1), "Outlier", adj=0)`, to place the label of an outlier close to the point.

**Exercise:** Make a scatterplot and mark a point as "My favourite".
```{r interactive, exercise = TRUE}
data(iris)

```


## Graphics Parameters
There are many graphing parameters, to change the look and feel of plots. For instance, the argument `pch = ` takes values between 0 and 25, it produces different plotting symbols (such as points). `lty =` stands for line type, `lwd = ` for line widths, `col = ` for the color of lines, texts, points and such, `cex = ` stand for character expansion.

**Exercise:** Make a scatterplot of two quantitative features of the Iris dataset. What happens if you set the colour to be like the `Species`?
```{r graphics-parameters, exercise = TRUE}
data(iris)


```


## Multiple Figure Environment
R allows you to create an $n$ by $m$ array of figures on a single page. This is usually done by setting, for instance, `mfcol = c(3, 2)` or `mfrow = c(2, 4)`. The first value is the number of columns, whereas the second is the number of rows. The difference is that in the first case the plots are filled column-wise and in the second they are filled row-wise.

**Exercise:** Make four plots in the same graph.
```{r multiple-figure, exercise = TRUE}
data(iris)

```

