---
title: "PCA: Example 1"
author: "Andrea Mascaretti"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```

# Example 1: PCA of simulated data

## Generating the dataset
We generate the data, after setting the seed. We are going to sample from a multivariate normal distribution.
We need to define $\mu$ and $\Sigma$, mean vector and covariance matrix. Moreover, we need to set $n$, the number of samples.

```{r 01-data, exercise = FALSE}
# Loading libraries ---------------------
library(mvtnorm)

# Setting the seed ----------------------
set.seed(26102016)

# Setting mean, variance, samples -------
mu  <-  c(1,2)
sig <-  matrix(data = c(1, 1, 1, 4), nrow = 2)
n   <-  100

# Sampling the data ---------------------
X <- rmvnorm(n, mu, sig)
```


## Analysis of the data


### Plot
We plot the data

```{r 01-plot, echo=TRUE, exercise=FALSE}
plot(X, asp=1, xlab='Var1', ylab='Var2', pch=19)
```

And add the sample mean

```{r 01-plot-sample-mean, echo=TRUE, exercise=FALSE}
plot(X, asp=1, xlab='Var1', ylab='Var2', pch=19)
points(colMeans(X)[1], colMeans(X)[2], col='red', pch=19,lwd=3)
```

We then plot the projection on the x and y axes and compute the variance.



### Axis X

```{r 01-plot-projection-x, echo=TRUE, exercise=FALSE}
plot(X, asp=1, xlab='Var1', ylab='Var2', pch=19)
points(colMeans(X)[1], colMeans(X)[2], col='red', pch=19,lwd=3)
abline(h=colMeans(X)[2], lty=2, col='grey')
points(X[,1], rep(colMeans(X)[2], n), col='red')
```

```{r 01-variance-x}
var(X[,1])
```



### Axis Y

```{r 01-plot-projection-y, echo=TRUE, exercise=FALSE}
plot(X, asp=1, xlab='Var1', ylab='Var2', pch=19)
points(colMeans(X)[1], colMeans(X)[2], col='red', pch=19,lwd=3)
abline(v=colMeans(X)[1], lty=2, col='grey')
points(rep(colMeans(X)[1], n), X[,2], col='red')
```

```{r 01-variance-y}
var(X[,2])
```


### Arbitrary direction
We can evaluate the variability of the dataset along any direction of the plane.

For instance, consider the direction with angle $\theta$ with the x axis:
$\theta = \frac{\pi}{6}$

```{r 01-plot-projecton-arbitrary, echo=TRUE, exercise=FALSE}
# Set theta ------------------------------
theta <- pi / 6

# Plot points ----------------------------
plot(X, asp=1, xlab='Var1', ylab='Var2', pch=19)
abline(a = colMeans(X)[2] - tan(theta)*colMeans(X)[1], b = tan(theta), lty=2)
a <- c(cos(theta), sin(theta))
proj30=(a)%*%(t(X)-colMeans(X))
points(colMeans(X)[1]+cos(theta)*proj30, 
       colMeans(X)[2]+sin(theta)*proj30, col='red')
```

Try to change the value of $\theta$ and see what happens!

```{r 01-plot-function, echo=FALSE, exercise=FALSE}
plot_direction <- function(X, theta) {
  tan_theta <- tan(theta)
  column_means <- colMeans(X)
  
  plot(X, asp = 1, xlab = 'Var1', ylab = 'Var2', pch = 19)
  abline(a = column_means[2] - tan_theta * column_means[1], b = tan_theta, lty = 2)
  a <- c(cos(theta), sin(theta))
  
  projection <- a %*% (t(X) - column_means)
  points(column_means[1] + cos(theta) * projection,
         column_means[2] + sin(theta) * projection,
         col='red')
}
```

```{r 01-plot-exercise, exercise=TRUE}
theta <- pi * 1/6
plot_direction(X = X, theta = theta)
```




### All the directions
We now compute the variance along all the directions.

```{r 01-theta, echo=TRUE, exercise=FALSE}
theta <- seq(from = 0, to = 2*pi, by = 2*pi/3600)
```

And we compute the variances.

```{r 01-variances-computation, exercise=FALSE}
N <- length(theta)
a <- matrix(data = c(cos(theta), sin(theta)), nrow = N, ncol = 2, byrow = FALSE)
Var <- apply(X = a, MARGIN = 1, FUN = function(x) cov(X %*% x))
```


### Analysis
We now want to see how the variances change according to the direction.

```{r 01-variance-directions, exercise=FALSE}
(max_var   <- max(Var))
(max_theta <- theta[which.max(Var)])
(max_a     <- c(cos(max_theta), sin(max_theta)))

(min_var   <- min(Var))
(min_theta <- theta[which.min(Var)])
(min_a     <- c(cos(min_theta), sin(min_theta)))
```



### Plotting the results

```{r 01-plotting-results, exercise=FALSE}
par(mfrow=c(1,2))
plot(X, asp=1, xlab='Var 1', ylab='Var 2',pch=20)
abline(a = colMeans(X)[2] - tan(max_theta)*colMeans(X)[1], b = tan(max_theta), lty = 4, col = 'navyblue', lwd = 2)
abline(a = colMeans(X)[2] - tan(min_theta)*colMeans(X)[1], b = tan(min_theta), lty = 4, col = 'red', lwd = 2)

plot(theta, Var, type = 'l', col='dark grey', lwd = 2,ylab='Variance')
points(max_theta, max_var, pch=16, col='navyblue')
points(min_theta, min_var, pch=16, col='red')
```



### Exact Computation
We now check what we have found against the actual values. We compute the loads (`loads`) and the variances (`var_pc`).

```{r 01-true-components, echo=TRUE, exercise=FALSE}
(loads <- eigen(cov(X))$vectors)
(var_pc <- eigen(cov(X))$values)
```

```{r 01-true-plot}
# Set the graph -----------------------------------------------
par(mfrow=c(1,2))

# Plot the points ---------------------------------------------
plot(X, asp=1, xlab='Var 1', ylab='Var 2',pch=20)

# Compute the column means of X -------------------------------
M <- colMeans(X)

# Plot directions of maximum and minimum variance -------------
abline(a = M[2] - loads[2,1]/loads[1,1]*M[1], b = loads[2,1]/loads[1,1], lty = 2, col = 'dark blue', lwd = 2)
abline(a = M[2] - loads[2,2]/loads[1,2]*M[1], b = loads[2,2]/loads[1,2], lty = 2, col = 'dark red', lwd = 2)

abline(a = M[2] - tan(max_theta)*M[1], b = tan(max_theta), lty = 4, col = 'navyblue', lwd = 2)
abline(a = M[2] - tan(min_theta)*M[1], b = tan(min_theta), lty = 4, col = 'red', lwd = 2)

# Plot the possible angles
plot(theta, Var, type = 'l', col='dark grey', lwd = 2,ylab='Variance')

# Plot the angles of maximum and minimum variances ------------
points(max_theta, max_var, pch=20, col='navyblue')
points(min_theta, min_var, pch=20, col='red')
points(atan(loads[2,1]/loads[1,1]), max_var, pch=3, col='dark blue')
points(atan(loads[2,2]/loads[1,2])+pi, min_var, pch=3, col='dark red')
```
