---
title: "dplyr: An Introduction "
author: "Andrea Mascaretti"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```


## Data Transformation
For this lab, we will follow [R for Data Science](https://r4ds.had.co.nz/transform.html#filter-rows-with-filter).
A first fundamental step for any statistical analysis is to explore the data and to derive some indicators to better understand the problem.
To see why this is important, we will work on a data frame containing all the flights that departed from New York City in 2013.
Moreover, we will rely on the [tidyverse](https://www.tidyverse.org/) suits of packages. We will, in particular, focus on the dplyr package.



### Loading the libraries
```{r loading-libraries}
library(tidyverse)
library(nycflights13)
```


### nycflights13
Let us have a look at the data.
```{r nycflights13-first-look}
flights
```

We have 19 columns. Each column contain some information, ranging from the year to the time of departure. Notice that for each column, we have a four letter abbreviation. Such abbreviation indicate the *type* of the variable.
The possible types are:

   - `int` for integers
   - `dbl` for doubles (real numbers)
   - `chr` for character vectors (strings)
   - `dttm` for date-times (a date and a time)
   - `lgl` for logical (`TRUE` or `FALSE` values)
   - `fctr` for factors, a type that R uses to represent categorical variables
   - `data` for dates.
   
   
## dplyr basics
There are five key operations that are possible by means of the dplyr package:

  - Filter observations
  - Reorder rows
  - Select a set of variables
  - Create new variables as functions of existing ones
  - Summarise variables

All the above functions can be used in conjunction with the `group_by` function, that changes the scope of the operations from the whole data frame to group-by-group batches.

The syntax is the same for every function:

  1. The first argument is the data frame (or *tibble* in the tidyverse parlance)
  1. The subsequent arguments describe the actions to take, using the variables name
  1. The result is a new tibble
  
This homogeneity allows for chaining different operations. Notice that **none** of the operations operates on the data frame passed as first argument.



### Filtering data
Filtering allows to select a subset of the rows (*i.e* the observations). Imagine, for instance, that we wish to select all the flights that departed on 16/03/2013.
```{r filter-data-example}
filter(flights, month == 3, day == 16)
```
We obtain a new tibble!

Comparison can be done by using the usual operators: `>`, `>=`, `<`, `<=`, `!=` (not equal) and `==` (equal).

Select all the flights that departed between the 12 and the 16 of May (included).
```{r filter-data-exercise, exercise=TRUE}

```
```{r filter-data-exercise-solution}
filter(flights, month == 5, day >= 12, day <= 16)
```

You can store the new data frame by using the `<-` operator, as usual.



### Comparing real numbers
Computers rely to approximation when dealing with real numbers. Take a look at the following:
```{r floating-point-example, exercise=FALSE}
sqrt(2) ^ 2 == 2
```
Weird, right? The imprecisions due to the implementation of numbers mean we cannot directly compare results of operations that are theoretically identical.
We resort to the following:
```{r floating-point-near-example, exercise=FALSE}
near(sqrt(2)^2, 2)
```



### Logical operators
Multiple arguments of the `filter` operations, are treated as clauses related by the "and" operation. We may, however, be interested in finding the flights that have departed on the 11th of March or May.
This is no problem. We have to use the "or" operator: `|` ("and" is `&`).
```{r filter-or-example, exercise = FALSE}
filter(flights, day == 11, month == 3 | month == 5)
```
Notice that you might be tempted to write `month == 3 | 5`. This **does not** work. R will compute the operations in order: `3 | 5` is equal to `TRUE`, `TRUE` is then equal to `1` (`FALSE` is `0`) and we will end up with flights that departed in January!

Another options, perhaps easier, is by means of the `%in%` operation.
```{r filter-in-example, exercise = FALSE}
filter(flights, day == 11, month %in% c(11, 12))
```
Remember that `c(11, 12)` is a vector of elements 11 and 12.



### Selecting variables
Sometimes we are interested in selecting only a number of variables.
This is easily done using the `select()` function!
```{r select-example, exercise=FALSE}
select(flights, year, month, day)
```
We can select all the columns (inclusive) between two.
```{r select-inclusive-example, exercise=FALSE}
select(flights, year:day)
```
Or, in a similar fashion, remove them.
```{r select-inclusive-remove-example, exercise=FALSE}
select(flights, -(year:day))
```

There is a number of helper functions that can be used with `select`:

   - `starts_with("xyz")`: names beginning by "xyz"
   - `ends_with("xyz")`: names ending by "xyz"
   - `contains("xyz")`: names containing "xyz"
   
Another option is to use the `rename()` function, that only changes the name of the variable.


### Adding new variables
Sometimes new variables have to be computed. To do this, we resort to the `mutate()` function.
```{r mutate-example}
flights_sml <- select(flights,
                      year:day,
                      ends_with("delay"),
                      distance,
                      air_time
)

mutate(flights_sml,
       gain = dep_delay - arr_delay,
       speed = distance / air_time * 60
)
```

Notice that you can refer to variables you have just created.
```{r mutate-referring-new-variables-example}
mutate(flights_sml,
       gain = dep_delay - arr_delay,
       hours = air_time / 60,
       gain_per_hour = gain / hours
)
```



#### Some creation functions
The key property of any functions used to create new variables is that it must be *vectorised*: it must take as input a vector and outputs a vector with the same number of elements.

 - Arithmetic operations such as `+`, `-`, `*`, ... work fine. Remember that they act element-wise!
 - Aggregate functions can be used as well. For instance, `x / sum(x)` gives the proportion of the total.
 
 
 
### Summarising variables
The last operation is `summarise()` (or `summarize()`).
```{r summarise-example}
summarise(flights, delay = mean(dep_delay, na.rm = TRUE))
```
Notice that we have put `na.rm = TRUE`. This means that we remove from the data frame all those observations for which we have no value. Why? Because otherwise we would not be able to compute the mean.
```{r summarise-na-example}
summarise(flights, delay = mean(dep_delay, na.rm = FALSE))
```


### Grouping by variables
`summarise()` is usually best employed on sub-groups of the data. For instance, we may be interested in understanding the *daily* delay, rather than that of 2013.
To this aim, we resort to the `group_by` function.
```{r group-by-example}
by_day <- group_by(flights, year, month, day)
summarise(by_day, delay = mean(dep_delay, na.rm = TRUE))
```

## Pipelines
### Piping operations
Let us say we are now interested in understanding whether there is a relationship between the distance of the destination and the average delay.
Let us do this together!

First of all, we need to group flights by the destination.
```{r distance-delay-group-exercise, exercise = TRUE}

```
```{r distance-delay-group-exercise-solution}
by_dest <- group_by(flights, dest)
```

Then, we have to compute the summary values we need. For each destination and arrival delay, we want to compute the average distance and average arrival delay. The variables are the `distance` and `arr_delay` variables. 
```{r distance-delay-summary-exercise, exercise = TRUE}
# Mind the NA data!

```
```{r distance-delay-summary-exercise-solution}
by_dest <- group_by(flights, dest)
delay <- summarise(by_dest,
                   dist = mean(distance, na.rm = TRUE),
                   delay = mean(arr_delay, na.rm = TRUE)
)
```



### Building the plot...
Impressive! Yet, we may be interested in computing the numbers of data we have for each destination.
How to do so? We use the `n()` function.
```{r distance-delay-summary-count-example}
by_dest <- group_by(flights, dest)
delay <- summarise(by_dest,
  count = n(),
  dist = mean(distance, na.rm = TRUE),
  delay = mean(arr_delay, na.rm = TRUE)
)
delay
```

In this way, we can filter out destinations that have too little data points. Let us say we set a threshold at 20. We also initially remove Honolulu for graphical reasons.
```{r filter-distance-delay-example}
delay <- filter(delay, count > 20, dest != "HNL")
```

We are ready to plot!
```{r distance-delay-plot-example}
ggplot(data = delay, mapping = aes(x = dist, y = delay)) +
  geom_point(mapping = aes(size = count), alpha = 1/3) +
  geom_smooth(se = FALSE)
```



### Finally pipelines!
So, we have done three operations:

  1. Grouped flights according to the destination
  1. Summarised to compute the average distance, the average delay and the number of observations per destination
  1. Filtered to remove points without enough observations and Honolulu, which is twice as far away as the next closest airport is.
  
To do so, we had to create intermediate data frames we are not really interested in. What we care about here are the *transformation*. To emphasise this, we can employ pipe: `%>%`.
Let's have a look at the following.
```{r distance-delay-pipe, exercise = FALSE}
delays <- flights %>%
  group_by(dest) %>%
  summarise(
    count = n(),
    dist = mean(distance, na.rm = TRUE),
    delay = mean(arr_delay, na.rm = TRUE)
  ) %>%
  filter(count > 20, dest != "HNL")
delays
```

In this way, we can focus only on the transformation. What a pipe does is basically to rewrite `f(x)` as `x %>% f()`, `g(f(x, y), z)` as `x %>% f(y) %>% g(z)` and so on.