---
title: "PCA: Exercise 1"
author: "Andrea Mascaretti"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(knitr)
library(learnr)
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```


# PCA: Exercise 1
Perform PCA on `USArrests` data.

```{r}
warning("princomp doesn't work")
```



## Exploratory Data Analysis

The first thing we need to do is to load the data.
```{r data-load}
data(USArrests)
```

Summarise the data frame.
```{r data-inspect-summary, exercise = TRUE}

```
```{r data-inspect-summary-solution}
summary(USArrests)
```

Plot the pairs.
```{r data-inspect-pairs, exercise = TRUE}

```
```{r data-inspect-pairs-solution}
plot(USArrests, pch = 19)
```


Plot boxplots of the data.
```{r data-inspect-boxplot, exercise = TRUE}
```
```{r data-inspect-boxplot-hint}
# boxplot(as.matrix(USArrests), col = "gold")
```



## PCA - Original Data

Compute the Principal Components.
```{r pca-original, exercise = TRUE}

```
```{r pca-original-solution}
pc.USArrests <- princomp(USArrests, scores = T)
pc.USArrests
```


### Loadings
Inspect the loadings.
```{r pca-original-loadings, exercise = TRUE}

```
```{r pca-original-loadins-solution}
pc.USArrests <- princomp(USArrests, scores = T)
load.rec <- pc.USArrests$loadings
load.rec
```


We report them graphically.
```{r pca-original-loadings-plot, exercise = TRUE}

```
```{r pca-original-loadings-plot-solution}
par(mar = c(1, 4, 0, 2), mfrow = c(4, 1))
for (i in 1:4) barplot(load.rec[, i], ylim = c(-1, 1))
```


### Explained Variance

We now consider the explained variance. Try to plot the explained variance per principal component.
```{r pca-original-explained-variance, exercise = TRUE}

```
```{r pca-original-explained-variance-solution}
layout(matrix(c(2, 3, 1, 3), 2, byrow = T))
barplot(pc.USArrests$sdev^2, las = 2, main = "Principal components", ylim = c(0, 7e3), ylab = "Variances")
abline(h = 1, col = "blue")
barplot(sapply(USArrests, sd)^2, las = 2, main = "Original variables", ylim = c(0, 7e3), ylab = "Variances")

abline(h = 1, col = "blue")
abline(h = 0.8, lty = 2, col = "blue")
pc.USArrests <- princomp(USArrests, scores = T)
axis(2, at = 0:10 / 10, labels = 0:10 / 10)
axis(1, at = 1:ncol(USArrests), labels = 1:ncol(USArrests), las = 2)
```



### Scores
Compute the scores and visualise the first ones.
```{r pca-original-scores, exercise = TRUE}

```
```{r pca-original-scores-solution}
scores.USArrests <- pc.USArrests$scores
head(scores.USArrests)
```


And plot the scores to get a visual grasp!
```{r pca-original-scores-plot, exercise = TRUE}

```
```{r pca-original-scores-plot-solution}
layout(matrix(c(1, 2), 2))
boxplot(USArrests, las = 2, col = "red", main = "Original variables")
scores.USArrests <- data.frame(scores.USArrests)
boxplot(scores.USArrests, las = 2, col = "red", main = "Scores")
```



Finally, make a biplot.
```{r pca-original-biplot, exercise = TRUE}

```
```{r pca-original-biplot-solution}
biplot(pc.USArrests, scale = 0, cex = .7)
```



## PCA - Standardised Data
### Data Preprocessing

We now consider the standardised data.
Standardise the data.
```{r pca-sd-data, exercise = TRUE}

```
```{r pca-sd-data-solution}
USArrests.sd <- scale(USArrests)
USArrests.sd <- data.frame(USArrests.sd)
```



It is always a good idea to start with some data visualisation. Let us make a boxplot!
```{r pca-sd-boxplot, exercise = TRUE}

```
```{r pca-sd-boxplot-solution}
boxplot(USArrests.sd, col = "gold")
```


### PCA
Let us compute the PCA.
```{r pca-sd-fit, exercise = TRUE}

```
```{r pca-sd-fit-solution}
pc.USArrests <- princomp(USArrests.sd, scores = T)
```


Inspect the loadings.
```{r pca-sd-loadings, exercise = TRUE}

```
```{r pca-sd-loadings-solution}
load.rec <- pc.USArrests$loadings
load.rec
```


We also wish to make a graphical representation of the loadings of the principal components.
```{r pca-sd-loadings-plot, exercise = TRUE}

```
```{r pca-sd-loadings-plot-solution}
par(mar = c(1, 4, 0, 2), mfrow = c(4, 1))
for (i in 1:4) barplot(load.rec[, i], ylim = c(-1, 1))
```




### Variance
Let us plot the explained variance per principal component.
```{r pca-sd-explained-variance, solution = TRUE}

```
```{r pca-sd-explained-variance-solution}
USArrests.sd <- scale(USArrests)
layout(matrix(c(2, 3, 1, 3), 2, byrow = T))
barplot(pc.USArrests$sdev^2, las = 2, main = "Principal components", ylim = c(0, 4), ylab = "Variances")
abline(h = 1, col = "blue")
barplot(sapply(USArrests.sd, sd)^2, las = 2, main = "Original variables", ylim = c(0, 4), ylab = "Variances")
plot(cumsum(pc.USArrests$sd^2) / sum(pc.USArrests$sd^2), type = "b", axes = F, xlab = "number of components", ylab = "contribution to the total variace", ylim = c(0, 1))
abline(h = 1, col = "blue")
abline(h = 0.8, lty = 2, col = "blue")
box()
axis(2, at = 0:10 / 10, labels = 0:10 / 10)
axis(1, at = 1:ncol(USArrests.sd), labels = 1:ncol(USArrests.sd), las = 2)
```



### Scores
Let us inspect the scores and inspect the first few.
```{r pca-sd-scores, exercise = TRUE}

```
```{r pca-sd-scores-solution}
scores.USArrests <- pc.USArrests$scores
head(scores.USArrests)
```



### Dispersion
We now wish to visualise the dispersion of the original data and the dispersion of the scores.
```{r pca-sd-dispersion, exercise = TRUE}

```
```{r pca-sd-dispersion-solution}
layout(matrix(c(1, 2), 2))
boxplot(USArrests.sd, las = 2, col = "red", main = "Original variables")
scores.USArrests <- data.frame(scores.USArrests)
boxplot(scores.USArrests, las = 2, col = "red", main = "Scores")
```


### Biplot
To conclude, let us make a biplot.
```{r pca-sd-biplot, exercise = TRUE}

```
```{r pca-sd-biplot-solution}
biplot(pc.USArrests, scale = 0, cex = .7)
```