---
title: "HT and ANOVA: Exercise 2"
author: "Andrea Mascaretti"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(knitr)
library(learnr)
knitr::opts_chunk$set(echo = TRUE, exercise = FALSE)
```


# Hypothesis test and ANOVA

## Introduction
We begin by importing the data set, building two datasets (female and male), containing both the biometrical and the social variables.


```{r data}
mydata <- read.table(file = "../data/cleaneddata.txt", header = TRUE, sep = ",")

mydata.gender <- mydata[, 1]
mydata.biosocial <- mydata[, 3:8]

t1 <- mydata.biosocial[mydata.gender == "Female", ]
t2 <- mydata.biosocial[mydata.gender == "Male", ]
```

We assess the dimension of the datasets.
```{r data-dim}
(p <- dim(t1)[2])
(n1 <- dim(t1)[1])
(n2 <- dim(t2)[1])
```

We also make a plot.
```{r data-plot}
plot(rbind(t1, t2), col = c(rep("red", n1), rep("black", n2)))
```



## Multivariate two-sample tests
### Adjusted two-sample Z tests for comparing the means (unpaired)

We set the level of significance,
```{r alpha}
alpha <- 0.05
```
we choose the null hypothesis,
```{r null-hypothesis}
delta.mean.H0 <- c(0, 0, 0, 0, 0, 0)
```
compute the three estimates of the differences between means,
```{r estimate-mean}
(delta.mean.est <- colMeans(t1) - colMeans(t2))
```
estimate the standard deviations (if unknown).
```{r estimate-variance}
(sd.est <- sqrt(apply(t1, 2, var) / n1 + apply(t2, 2, var) / n2))
```
We then proceed to compute the Z test statistics,
```{r z-test-stats}
(zstat <- (delta.mean.est - delta.mean.H0) / sd.est)
```
adjust the level of significance (*e.g.* Bonferroni adjustment),
```{r adjusted-alpha}
(alpha.adj <- alpha / p)
```
compute the boundaries of the rejection region,
```{r boundaries}
(cfr.z <- qnorm(1 - alpha.adj / 2))
```
check if this leads to the rejection of $H_0$ or not,
```{r H0-rejection}
(abs(zstat) > cfr.z)
```
and finally compute the p-value
```{r p-value}
(pvalue <- ifelse(zstat >= 0, (1 - pnorm(zstat)) * 2, pnorm(zstat) * 2))
```
and the adjusted p-value.
```{r adjusted-p-value}
(pvalue.adj <- pmin(pvalue * p, 1))
```

We also inspect the adjust two-sample Z intervals for comparing *unpaired* means.
```{r two}
CI <- cbind(
  inf = delta.mean.est - sd.est * qnorm(1 - alpha.adj / 2),
  center = delta.mean.est,
  sup = delta.mean.est + sd.est * qnorm(1 - alpha.adj / 2)
)
kable(CI)
```







## Global Testing
We now proceed by applying a global testing approach: Hotelling's test.


We select the level of significance,
```{r ho-alpha}
(alpha <- 0.05)
```
choose the null hypothesis (a vector in $\mathbb{R}^3$),
```{r ho-H0}
(delta.mean.H0 <- c(0, 0, 0, 0, 0, 0))
```
compute the three estimate of the three differences between means (a vector in $\mathbb{R}^3$),
```{r ho-means}
(delta.mean.est <- colMeans(t1) - colMeans(t2))
```
estimate the covariance matrix (a matrx in $\mathbb{R}^{3 \times 3}$),
```{r ho-covariance}
(cov.est <- cov(t1) / n1 + cov(t2) / n2)
```
compute the $\chi^2$ test statistic,
```{r ho-chi-square}
(chi2stat <- (delta.mean.est - delta.mean.H0) %*% solve(cov.est) %*% (delta.mean.est - delta.mean.H0))
```
compute the boundary of the elliptical rejection region,
```{r ho-boundary}
(cfr.chi2 <- qchisq(1 - alpha, p))
```
check if this leads to the rejection of $H_0$ or not,
```{r ho-reject-H0}
(chi2stat > cfr.chi2)
```
we then compute the p-value.
```{r ho-p-value}
(pvalue <- 1 - pchisq(chi2stat, p))
```


We inspect $\chi^2$ confidence regions for comparing the *unpaired* means (which can't be visualised whenever $p > 3$)-
```{r ho-confidence-region}
CI.sim <- cbind(
  inf = delta.mean.est - sd.est * sqrt(qchisq(1 - alpha, p)),
  center = delta.mean.est,
  sup = delta.mean.est + sd.est * sqrt(qchisq(1 - alpha, p))
)
kable(CI.sim)
```







## ANOVA
We now turn our attention to the ANOVA.

We load the data.
```{r iris-data}
iris
summary(iris)
```

We build the four vector of responses:

    - `Sepal.Length`
    - `Sepal.Width`
    - `Petal.Length`
    - `Petal.Width`
    
and the vector of treatment levels:

    - `Species`
    
```{r iris-variable}
response1 <- iris[, 1]
response2 <- iris[, 2]
response3 <- iris[, 1]
response4 <- iris[, 4]

treatment <- iris[, 5]
```



## Dimensions
We then check the sample size,
```{r iris-sample-size}
(n <- length(response1))
```
the group sample size,
```{r iris-group-sample-size}
(ng <- table(treatment))
```
the number of treatments,
```{r iris-no-treatments}
(treatment.levels <- levels(treatment))
(g <- length(treatment.levels))
```
the response to explore.
```{r iris-response}
(response <- response4)
```
We plot the treatment groups.
```{r iris-treatment}
(boxplot(response ~ treatment))
```





## Test
We ought to test the equality of the six means of the treatment groups:


   - $H_0: \tau_{1} = \tau_2 = \tau_3 = \tau_4 = \tau_5 = \tau_6 = 0$
   - $H_1: H_0^c$
   
We compute the estimates of the $g$ means under the null hypothesis.
```{r H0-g-means}
rep(mean(response), g)
```


We compute the estimates of the $g$ means under the alternative hypothesis. Note that these are valide (albeit sub-optimal) estimates even under the null hypothesis.
```{r H1-g-means}
tapply(response, treatment, mean)
```



### ANOVA
We perform the ANOVA decomposition.
```{r iris-anova}
fit <- aov(response ~ treatment)
summary(fit)
```



We now perform pair-wise comparisons (*e.g.* Bonferroni). We need to make $\frac{g \left(g - 1\right)}{2}$ comparisons.
```{r iris-pairwise-comparison}
mean.est.matrix <- matrix(NA, nrow = g, ncol = g, dimnames = list(treatment.levels, treatment.levels))
zstat.matrix <- matrix(NA, nrow = g, ncol = g, dimnames = list(treatment.levels, treatment.levels))
pvalue.matrix <- matrix(NA, nrow = g, ncol = g, dimnames = list(treatment.levels, treatment.levels))
pvalue.adj.matrix <- matrix(NA, nrow = g, ncol = g, dimnames = list(treatment.levels, treatment.levels))
for (i in 1:(g - 1)) {
  for (j in (i + 1):g) {
    test.ij <- t.test(response[treatment == treatment.levels[i]], response[treatment == treatment.levels[j]])
    mean.est.matrix[i, j] <- diff(test.ij$estimate)
    zstat.matrix[i, j] <- test.ij$statistic
    pvalue.matrix[i, j] <- test.ij$p.value
    pvalue.adj.matrix[i, j] <- min(test.ij$p.value * (g * (g - 1) / 2), 1)
  }
}
```



We then plot the pairwise estimated differences betwen the means,
```{r iris-pw-mean-diff}
mean.est.matrix
```
the pairwise estimated test statistics,
```{r iris-pw-test-stats}
zstat.matrix
```
the pairwise p-values
```{r iris-pw-p-values}
pvalue.matrix
```
and the pairwise adjusted p-values.
```{r iris-pw-adj-p-values}
pvalue.adj.matrix * 4
pvalue.adj.matrix * 4 < 0.05
```





## Two-way ANOVA
We load the data.
```{r mineral-data}
npk
summary(npk)
```


We have:


     - Response: yield [pounds/plot]
     - Factor 1: Block (1, 2, 3, 4, 5, 6)
     - Factor 2: Nitrogen (0: No, 1: Yes)
     - Factor 3: Phosphate (0: No, 1: Yes)
     - Factor 4: Potassium (0: No, 1: Yes)

We create factor-specific boxplots of yields.
```{r mineral-boxplots}
boxplot(npk[, "yield"] ~ npk[, "block"])
boxplot(npk[, "yield"] ~ npk[, "N"])
boxplot(npk[, "yield"] ~ npk[, "P"])
boxplot(npk[, "yield"] ~ npk[, "K"])
```



We fit an ANOVA model which is additive with respect to blocks, while it admits pairwise interactions between $N$, $P$ and $K$.
```{r mineral-anova}
fit <- aov(yield ~ block + N + P + K + N:P + N:K + P:K, npk)
```

We analyse the results:
```{r mineral-anova-analysis}
summary.aov(fit)
```


We select a simplified model.
```{r mineral-anova-simplified}
fit1 <- aov(yield ~ block + N + K, npk)
summary(fit1)
```

