---
title: "HT and ANOVA: Exercise 1"
author: "Andrea Mascaretti"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(knitr)
library(learnr)
knitr::opts_chunk$set(echo = TRUE, exercise = FALSE)
```

# Hypothesis Testing and Anova

## Two-Sample Z tests for comparing the means (unpaired)
### Data Import
We import the data
```{r zun-data}
t1 <- read.table(file = "../data/barcellona.txt", header = TRUE)
t2 <- read.table(file = "../data/milano.txt", header = TRUE)

p <- dim(t1)[2]
n1 <- dim(t1)[1]
n2 <- dim(t2)[1]
```

and plot them.
```{r zun-data-plot}
plot(rbind(t1, t2), col = c(rep("red", n1), rep("black", n2)))
```




### Hypothesis
We select the level of significance
```{r zun-alpha}
alpha <- 0.01
```


and the null hypothesis.
```{r zun-h0}
delta.mean.H0 <- 0
```

We then compute the estimate of the difference between the means
```{r zun-estimate-mean}
(delta.mean.est <- mean(t1[, "Temperatura"]) - mean(t2[, "Temperatura"]))
```
and estimate the standard deviation (if it is not known)
```{r zun-estimate-sd}
(sd.est <- sqrt(var(t1[, "Temperatura"]) / n1 + var(t2[, "Temperatura"]) / n2))
```
We compute the Z-test statistic
```{r zun-z}
(zstat <- (delta.mean.est - delta.mean.H0) / sd.est)
```
and compute the boundaries of the rejection region.
```{r zun-rejection-region}
(cfr.z <- qnorm(1 - alpha / 2))
```
We check if this leads to the rejection of $H_{0}$ or not
```{r zun-h0-accepted}
abs(zstat) > cfr.z
```
and compute the p-value.
```{r zun-pvalue}
(pvalue <- ifelse(zstat >= 0, (1 - pnorm(zstat)) * 2, pnorm(zstat) * 2))
```


We finally compute the two-sample Z intervals for comparing the means (unpaired)
```{r zun-z-intervals}
CI <- c(
  inf = delta.mean.est - sd.est * qnorm(1 - alpha / 2),
  center = delta.mean.est,
  sup = delta.mean.est + sd.est * qnorm(1 - alpha / 2)
)
kable(CI)
```





## Multivariate two-sample tests for comparing the means (unpaired)

We consider a component-wise testing approach: an adjusted two-sample Z test for comparing the means (unpaired).

We select the level of signficance,
```{r muzun-alpha}
alpha <- 0.01
```
the null hypothesism
```{r muzun-h0}
delta.mean.H0 <- c(0, 0, 0)
```
we compute three estimates of the three differences of the means.
```{r muzun-estimate-mean}
(delta.mean.est <- colMeans(t1) - colMeans(t2))
```
After that, we estiamte the standard deviations of the three estimates (if unkwonw),
```{r muzun-estimate-sd}
(sd.est <- sqrt(apply(t1, 2, var) / n1 + apply(t2, 2, var) / n2))
```
we compute the Z-test statistics,
```{r muzun-z-test}
(zstat <- (delta.mean.est - delta.mean.H0) / sd.est)
```
adjust the level of signficance (*e.g.* Bonferroni adjustment);
```{r muzun-bonferroni}
(alpha.adj <- alpha / p)
```
we compute the boundaries of the rejection regions,
```{r muzun-boundaries}
(cfr.z <- qnorm(1 - alpha.adj / 2))
```
check if the boundaries lead to the rejection of $H_0$ or not,
```{r muzun-h0-boundaries}
abs(zstat) > cfr.z
```
compute the p-value,
```{r muzun-pvalue}
(pvalue <- ifelse(zstat >= 0, (1 - pnorm(zstat)) * 2, pnorm(zstat) * 2))
```
the adjusted p-value,
```{r muzun-adjusted-pvalue}
(pvalue.adj <- pmin(pvalue * p, 1))
```
and finally show the adjusted two-sample Z intervals for comparing the means (unpaird) (*e.g.* Bonferroni intervals)
```{r muzun-z-intervals}
CI <- cbind(
  inf = delta.mean.est - sd.est * qnorm(1 - alpha.adj / 2),
  center = delta.mean.est,
  sup = delta.mean.est + sd.est * qnorm(1 - alpha.adj / 2)
)
kable(CI)
```





## Global Testing Approach (Hotelling's test)
We now approach the issue by using a global testing approach, namely Hotelling's test.

We select the level of signficance,
```{r ho-alpha}
alpha <- 0.01
```
choose the null hypothesis (a vector in $\mathbb{R}^3$).
```{r ho-null}
delta.mean.H0 <- c(0, 0, 0)
```
compute three estimates of the the three differences of the means (again, a vector in $\mathbb{R}^3$),
```{r ho-mean-est}
(delta.mean.est <- colMeans(t1) - colMeans(t2))
```
we estimate the covariance matrix (a matrix in $\mathbb{R}^{3 \times 3}$).
```{r ho-cov-est}
(cov.est <- cov(t1) / n1 + cov(t2) / n2)
```
We then compute the $\chi^{2}$-test statistic,
```{r ho-chi-square}
(chi2stat <- (delta.mean.est - delta.mean.H0) %*% solve(cov.est) %*% (delta.mean.est - delta.mean.H0))
```
compute the boundary of the elliptical rejection region,
```{r ho-boundaries}
(cfr.chi2 <- qchisq(1 - alpha, p))
```
check if this leads to the rejection of $H_{0}$ or not
```{r ho-H0-rejection}
chi2stat > cfr.chi2
```
and compute the p-value.
```{r ho-pvalue}
(pvalue <- 1 - pchisq(chi2stat, p))
```
Moreover, we show the $\chi^2$ confide:nce region for comparing the means (unpaired). They can't be visualised whenever $p > 3$.
We are interested in checking the shades along the three axes (*i.e.*, simultaneous confidence intervals)
```{r ho-shades}
CI.sim <- cbind(
  inf = delta.mean.est - sd.est * sqrt(qchisq(1 - alpha, p)),
  center = delta.mean.est,
  sup = delta.mean.est + sd.est * sqrt(qchisq(1 - alpha, p))
)
kable(CI.sim)
```







## One-Way ANOVA
### Introduction
We load the data. We have $p = 1$ and $g = 6$.
```{r chick-data}
chickwts
summary(chickwts)
```


We build the vector of responses (*e.g.* weight) and the vector of treatment levels (*e.g.* feed type)
```{r chick-data-set}
response <- chickwts[, "weight"]
treatment <- chickwts[, "feed"]
```


We compute the sample size,
```{r chick-sample}
n <- length(response)
```
the treatment group sample sizes
```{r chick-sample-groups}
ng <- table(treatment)
```
and the number of treatments.
```{r chick-number-treatments}
treatment.levels <- levels(treatment)
g <- length(treatment.levels)
```



We also plot the treatment groups.
```{r chick-groups-boxplot}
boxplot(response ~ treatment)
```






### Computing ANOVA

We wish to test for the equality of the $g$ means of the treatment groups: 


    - $H_{0}: \, \tau_{1} = \tau_{2} = \tau_{3} = \tau_{4} = 0$
    - $H_{1}: \, H_{0}^c$
    

We compute the estimates of the $g$ means under the null hypothesis,
```{r chick-mean-estimate-null}
rep(mean(response), g)
```

and we compute the estimates of the six means under the alternative hypothesis.
*Note that these are valid (albeit sub-optimal) estimates even under the null hypothesis.*
```{r chick-mean-estimate-alternative}
tapply(response, treatment, mean)
```



We perform the ANOVA decomposition.
```{r chick-anova-fit}
fit <- aov(response ~ treatment)
summary(fit)
```


And we perform pairwise comparision (*e.g.* Bonferroni). We need to make $g \dot \frac{g - 1}{2}$ comparisons.
```{r chick-comparison}
mean.est.matrix <- matrix(NA, nrow = g, ncol = g, dimnames = list(treatment.levels, treatment.levels))
zstat.matrix <- matrix(NA, nrow = g, ncol = g, dimnames = list(treatment.levels, treatment.levels))
pvalue.matrix <- matrix(NA, nrow = g, ncol = g, dimnames = list(treatment.levels, treatment.levels))
pvalue.adj.matrix <- matrix(NA, nrow = g, ncol = g, dimnames = list(treatment.levels, treatment.levels))
for (i in 1:(g - 1)) {
  for (j in (i + 1):g) {
    test.ij <- t.test(response[treatment == treatment.levels[i]], response[treatment == treatment.levels[j]])
    mean.est.matrix[i, j] <- diff(test.ij$estimate)
    zstat.matrix[i, j] <- test.ij$statistic
    pvalue.matrix[i, j] <- test.ij$p.value
    pvalue.adj.matrix[i, j] <- min(test.ij$p.value * (g * (g - 1) / 2), 1)
  }
}
```



We now plot the pairwise estimated differences of the means,
```{r chick-means}
(mean.est.matrix)
```
the pairwise estimated test statistics,
```{r chick-test}
(zstat.matrix)
```
the pairwise p-values
```{r chick-pvalues}
(pvalue.matrix)
```
and the pairwise adjusted p-values.
```{r chick-adjust-pvalues}
(pvalue.adj.matrix)
pvalue.adj.matrix < 0.05
```






## Two-way ANOVA
We compute a two-way ANOVA for the following problem:

    - Respone: Distance per fuel litre [km/l]
    - Factor 1: Gasoline Supplier (0: Esso, 1: Shell)
    - Factor 2: Gasoline Type (0: 95, 1: 98)
    
We set the variables.
```{r gas-data}
km <- c(18.7, 16.8, 20.1, 22.4, 14.0, 15.2, 22.0, 23.3)
gaz.supplier <- factor(c("Esso", "Esso", "Esso", "Esso", "Shell", "Shell", "Shell", "Shell"))
gaz.type <- factor(c("95", "95", "98", "98", "95", "95", "98", "98"))
gaz.supplier.type <- factor(c("Esso95", "Esso95", "Esso98", "Esso98", "Shell95", "Shell95", "Shell98", "Shell98"))
```



We estimate the means under the hypothesis of no effect of supplier nor type.
```{r gas-no-effect}
(M <- mean(km))
```
We estimate the means under the hypothesis of no effect of type only.
```{r gas-no-type}
(Md.supplier <- tapply(km, gaz.supplier, mean))
```
We estimate the means under the hypothesis of no effect of supplier only.
```{r gas-no-supplier}
(M.type <- tapply(km, gaz.type, mean))
```
We estimate the means under the hypothesis of an effect of both supplier and type.
```{r gas-both}
(M.supplier.type <- tapply(km, gaz.supplier.type, mean))
```


And perform the analysis.
```{r gas-anova}
summary.aov(aov(km ~ gaz.supplier + gaz.type + gaz.supplier:gaz.type))

summary.aov(aov(km ~ gaz.supplier + gaz.type))

summary.aov(aov(km ~ gaz.type))
```

