---
title: "Univariate Descriptive"
author: "Andrea Mascaretti"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(knitr)
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, exercise = FALSE)
```


# Univariate Descriptive
## Example 1
### Introduction

First of all, we load the required libraries.
```{r load-libraries}
library(MASS)
library(plotrix)
library(lattice)
library(moments)
library(tidyverse)
```


Our goal is to analyse the data contained within the `magnesium.txt` file.
The file contains the measurements of the quantity of magnesium [mmol/l] in the blood of 140 healthy subjects.

Let us load the data.
```{r mag-load-data}
data("magnesio")
attach(magnesio)
```

Let us double-check the length of the dataset.
```{r mag-length-data}
(n <- length(magnesio$Magnesio))
```





### Location and shape indices
We compute some location and shape indices.
Namely, the minimum,
```{r mag-min}
min(Magnesio)
```

the maximum,
```{r mag-max}
max(Magnesio)
```

the mean,
```{r mag-mean}
mean(Magnesio)
```

the standard deviation,
```{r mag-sd}
sqrt(var(Magnesio))
sd(Magnesio)
```

the range,
```{r mag-range}
range(Magnesio)
```

the quartiles and interquartile range,
```{r mag-quartiles}
quantile(Magnesio, 0.25)
quantile(Magnesio, 0.5)
quantile(Magnesio, 0.75)
quantile(Magnesio) # if no specification are provided, all the quartiles are computed

IQR(Magnesio)
```

(moments can also be computed like this:)
```{r}
(mom <- all.moments(magnesio, central = TRUE, order.max = 4))
```

skewness (notice that the skewnes of symmetric data is $0$),
```{r mag-skewness}
skewness(magnesio)
(sum((Magnesio - mean(Magnesio))^3) / n) / (sum((Magnesio - mean(Magnesio))^2) / n)^(3 / 2)
mom[3 + 1] / mom[2 + 1]^(3 / 2)
```



kurtosis (and notice that the kurtosis of the normal distribution is $3$.
```{r mag-curtosis}
kurtosis(magnesio)
(sum((Magnesio - mean(Magnesio))^4) / n) / (sum((Magnesio - mean(Magnesio))^2) / n)^2
mom[4 + 1] / mom[2 + 1]^2
```


We detach the data.
```{r mag-detach}
detach(magnesio)
```






### Comparison with Normal
Note what happens to the skeweness and Kurtosis indices when distributions are symmetrical or not.
Since skewness is a measure of symmetry (more precisely, of the lack of symmetry), data sets with high kurtosis tend to have heavy tails, or outliers.
Since Kurtosis is a measure of whether the data are heavy-tailed or light-tailed relative to a normal distribution, data sets with low kurtosis tend to have light tails, or lack of outliers.


We now check this against a Normal distrubtion (which is symmetric)
```{r normal}
dat <- rnorm(100000, mean = 0, sd = 1)
summary(dat)
```

The skewness is almost $0$, like in all symmetric distributions.
```{r normal-skewness}
skewness(dat)
```

Kurtosis is close to $3$: light tails and no tendency to outlier with respect to the Normal distribution.
```{r normal-kurtosis}
kurtosis(dat)
```




We make the same analysis simulating an exponential distribution. We set the rate $\lambda = 2$. The theoretical mean is equal to $\frac{1}{\lambda}$.


```{r exponential}
dat <- rexp(100000, rate = 2) # the theoretical mean is equal to 1/rate
summary(dat)
```


We make an histogram.
```{r exponential-histogram}
hist(dat)
```

The positive skewness means that the right tail is long with respect to the left tail.
```{r}
skewness(dat)
```

The high kurtosis implies a heavy tail and a tendency to outliers with respect to the Normal distribution.
```{r}
kurtosis(dat)
```






## Example 2
We now consider the `cavendish.txt` data file, from [Stigler](https://projecteuclid.org/euclid.aos/1176343997). It consists of $29$ measurements of earth density, carried out by Henry Cavendish in 1798.
The goal is to analyse Cavendish data and detect any outlier.

```{r caven-data}
data(cav)
attach(caven)
```


### Location and shape indices
We compute some location and shape indices.
Namely, the minimum,
```{r caven-min}
min(Density)
```

the maximum,
```{r caven-max}
max(Density)
```

the mean,
```{r caven-mean}
mean(Density)
```

the standard deviation,
```{r caven-sd}
sqrt(var(Density))
sd(Density)
```

the range,
```{r caven-range}
range(Density)
```

the quartiles and interquartile range,
```{r caven-quartiles}
(q1 <- quantile(Density, 0.25))
quantile(Density, 0.5)
(q3 <- quantile(Density, 0.75))
quantile(Density)

(iqr <- IQR(Density))
```

(moments can also be computed like this:)
```{r caven-moments}
(mom <- all.moments(Density, central = TRUE, order.max = 4))
```

skewness (notice that the skewnes of symmetric data is $0$),
```{r caven-skewness}
skewness(Density)
(sum((Density - mean(Density))^3) / n) / (sum((Density - mean(Density))^2) / n)^(3 / 2)
mom[3 + 1] / mom[2 + 1]^(3 / 2)
```



kurtosis (and notice that the kurtosis of the normal distribution is $3$.
```{r caven-curtosis}
kurtosis(Density)
(sum((Density - mean(Density))^4) / n) / (sum((Density - mean(Density))^2) / n)^2
mom[4 + 1] / mom[2 + 1]^2
```




We make some graphical representations of the continuous variable `Density`.
```{r caven-histogram}
histogram <- hist(Density, plot = TRUE, breaks = seq(4, 6, .25))
```

```{r caven-boxplot}
boxplot(Density, horizontal = FALSE, main = "Boxplot of Density", ylab = "earth density")
```

Notice the plot below the lower whisker.





### Outliers Identification
We try to identify potential outliers.
```{r caven-outliers}
Density[which(Density < q1 - 1.5 * iqr)]
Density[which(Density > q3 + 1.5 * iqr)]
```

We find the point below the whisker!
We remove the outlier.


```{r new-caven-data}
new_density <- Density[-which(Density < q1 - 1.5 * iqr)]
new_density
```


Check the lengths!
```{r new-caven-length}
length(Density)
length(new_density)
```





### Comparison among location and shape indices after removing the outlier.

We compare location and shape indices *before* and *after* the removal of the outlier.
Robust indices do not change a lot.
```{r new-caven-robust-compare}
knitr::kable(rbind(quantile(Density), quantile(new_density)))
```

The mean and (in particular) the standard deviation are more heavily affected.
```{r new-caven-compare}
knitr::kable(rbind(cbind(mean(Density), sd(Density)), cbind(mean(new_density), sd(new_density))), col.names = c("Mean", "Standard Deviation"))
```


We also make a boxplot of the data without the outlier.
```{r new-caven-boxplot}
boxplot(new_density, horizontal = FALSE, main = "Boxplot Density", ylab = "earth density")
```

We are finally ready to detach the data set.
```{r caven-detach}
detach(caven)
```





## Exercise I
Repeat the previous analysis on the dataset `serum.txt`:

    * Compute suitable location and dispersion indices
    * Give a graphical representation of the variable
    * Identify (*if applicable*) any outliers and compare the results with and without removing them from the dataset.
    
Data are taken from [[Bland]](https://global.oup.com/academic/product/an-introduction-to-medical-statistics-9780199589920?cc=it&lang=en&). They contain the triglyceride concentration in blood serum [mmol/l] of the umbilical cord in 282 subjects.
 
```{r exercise-serum}
data(serum)
```
 





## Example 3

We now turn our attention to a third dataset, as found in [Mackowiak and Wasserman](https://www.ncbi.nlm.nih.gov/pubmed/1302471).
The dataset comprises 130 tri-variate observations:

    1. Temperature (*temperature*) [°F]
    1. Sesso (*gender*) (U: Male, D: Female)
    1. Freq cardiaca (*cardiac frequency*) [pulse/min]
    
The goals of the analysis are to:

    1. Describe the variable `Temperatura` through suitable indices.
    1. Plot it and interpret it.
    1. Compare, both graphically and quantitatively) the distribution of male and female temperatures.
    


```{r temp-data}
data(temp)
dim(temp)
dimnames(temp)
temp
attach(temp)

n <- length(Temperatura)
```


The mean is:
```{r temp-mean}
mean(Temperatura)
```

The variance is:
```{r temp-var}
var(Temperatura)
```


The population variance is obtained donig $$\frac{n - 1}{n} \mathrm{Var}\left(\mathrm{Temperatura}\right).$$

The standard deviation is:
```{r temp-sd}
sd(Temperatura)
```

The population standard deviation is obtained donig $$\frac{n - 1}{n} \mathrm{sd}\left(\mathrm{Temperatura}\right).$$



The minimum and the maximum are:
```{r temp-max-min}
min(Temperatura)
max(Temperatura)
```


The range:
```{r temp-range}
range(Temperatura)
diff(range(Temperatura))
```



The quantiles:
```{r temp-quantiles}
quantile(Temperatura)
Q1 <- quantile(Temperatura, 0.25)
Q3 <- quantile(Temperatura, 0.75)

IQR(Temperatura)
```





### Graphical Representations
We make a histogram
```{r temp-hist}
hist(Temperatura, prob = TRUE, main = "Histogram", xlab = "body temperature [°F]", ylab = "density")
```


and a boxplot.
```{r temp-boxplot}
boxplot(Temperatura, horizontal = FALSE, main = "Boxplot", ylab = "body temperature [°F]", ylim = c(96, 101))
```




### Comparison

To aid us with the analysis, we use the `tapply()` function.
It takes three main arguments:

     * the variable of interest
     * the stratifying categorical variable
     * the operation to be applied to the first argument, stratified according to the categories introduced by the second argument.

We compute the mean.
```{r temp-gender-mean}
tapply(Temperatura, Sesso, mean)
```

**Exercise**
What do the following mean:

  * `tapply(Temperatura, Sesso, var)`
  * `tapply(Temperatura, Sesso, sd)`
  * `tapply(Temperatura, Sesso, min)`
  * `tapply(Temperatura, Sesso, max)`



We now compute the range.
```{r temp-gender-range}
tapply(Temperatura, Sesso, range)
```


To compute the range within the categories, we resort to the `diff()` function.
```{r temp-gender-range-diff}
diff(tapply(Temperatura, Sesso, range)$D)
diff(tapply(Temperatura, Sesso, range)$U)
```


Let us investigate the quantiles. We print the IQR of women and men, respectively, noting that men's IQE is wider than women's IQR.
```{r temp-gender-quantile}
(Q <- tapply(Temperatura, Sesso, quantile))
Q1 <- c(Q$D[2], Q$U[2])
Q3 <- c(Q$D[4], Q$U[4])
Q3 - Q1
```


**Exercise**
Compute the $90^{\mathrm{th}}$ quantile for the variable `Temperatura` using the `tapply()` command.
```{r temp-gender-quantile-90, exercise = TRUE}

```
```{r temp-gender-quantile-90-solution}
Q_90 <- tapply(Temperatura, Sesso, quantile, probs = 0.9)
Q_90
```








### Graphical Representation
We now make some histrograms.




### Flank Histograms
```{r temp-gender-flank-histo}
par(mfrow = c(1, 2)) # to make partition of the graph window
hist(Temperatura[Sesso == "D"],
  prob = TRUE, main = "Histogram Women", xlab = "body temperature [°F]",
  ylab = "Density", col = "pink", xlim = range(Temperatura), breaks = seq(96, 101, .25), ylim = c(0, 1)
)
hist(Temperatura[Sesso == "U"],
  prob = TRUE, main = "Histogram Men", xlab = "body temperature [°F]",
  ylab = "Density", col = "lightblue", xlim = range(Temperatura), breaks = seq(96, 101, .25), ylim = c(0, 1)
)
```



### Overlap Histograms
```{r temp-gender-overlap}
par(mfrow = c(1, 2))
hist(Temperatura[Sesso == "D"],
  prob = TRUE, main = "Histogram Men Women", xlab = "body temperature [°F]",
  ylab = "Density", col = "pink", xlim = range(Temperatura), breaks = seq(96, 101, .25), ylim = c(0, 1)
)
hist(Temperatura[Sesso == "U"],
  prob = TRUE, xlab = "body temperature [°F]",
  ylab = "Densit?", col = "lightblue", xlim = range(Temperatura), breaks = seq(96, 101, .25), ylim = c(0, 1), add = TRUE
)
hist(Temperatura[Sesso == "U"],
  prob = TRUE, main = "Histogram Women Men", xlab = "body temperature [°F]",
  ylab = "Density", col = "lightblue", xlim = range(Temperatura), breaks = seq(96, 101, .25), ylim = c(0, 1)
)
hist(Temperatura[Sesso == "D"],
  prob = TRUE, xlab = "body temperature [°F]",
  ylab = "Densit?", col = "pink", xlim = range(Temperatura), breaks = seq(96, 101, .25), ylim = c(0, 1), add = TRUE
)
```



### ggplot
```{r temp-gender-ggplot2}
temp %>%
        tibble() %>%
        ggplot() +
        geom_density(mapping = aes(Temperatura, fill = Sesso), alpha = 0.2)
```



### BOXPLOTS

We now turn our attention to boxplots.




### Basic Boxplot
```{r temp-gender-boxplot-basic}
boxplot(Temperatura ~ Sesso,
  data = temp, horizontal = FALSE, main = "Boxplot", names = c("Donne", "Uomini"), col = c("pink", "lightblue"),
  ylab = "body temperature [°F]", ylim = c(94, 102)
)
```






### ggplot Boxplot
```{r temp-gender-boxplot-ggplot}
temp %>%
        tibble() %>%
        ggplot() +
        geom_boxplot(mapping = aes(x = Sesso, y = Temperatura, fill = Sesso)) +
        scale_x_discrete(name = "Gender") +
        scale_y_continuous(name = "Temperature [°F]")

```



### Conclusion
To conclude, we detach the data.
```{r temp-detac}
detach(temp)
```
