---
title: "Clustering - Iris"
author: "Andrea Mascaretti"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE, exercise = FALSE)
```


# Hierarchical Agglomerative Clustering

We firstly load the libraries.

```{r load-library}
library(RColorBrewer)
```



## Loading the data

We load the data.

```{r load-data}
df <- iris[, 1:4]
```


## Computing distances

We also compute the pairwise distances and plot the results.

```{r distances}
iris_e <- dist(x = df, method = "euclidean")
```

```{r distances-plot}
image(1:150, 1:150, as.matrix(iris_e), main='Euclidean pairwise distances', asp=1, xlab = "", ylab = "")
```


## Dendogram

We build dendograms considering different linkages.

```{r dendograms}
iris_es <- hclust(d = iris_e, method='single')
iris_ea <- hclust(d = iris_e, method='average')
iris_ec <- hclust(d = iris_e, method='complete')
```


We plot the results we have obtained.

```{r dendograms-plot}
par(mfrow = c(1, 3))
plot(iris_es, main='Euclidean Single', hang=-0.1, xlab='', labels=FALSE, cex=0.6, sub='')
plot(iris_ec, main='Euclidean Complete', hang=-0.1, xlab='', labels=FALSE, cex=0.6, sub='')
plot(iris_ea, main='Euclidean Average', hang=-0.1, xlab='', labels=FALSE, cex=0.6, sub='')
```



## Cophenetic Matrices

Moreover, we compute the cophenetic matrices.

```{r cophenetic-matrices}
coph_es <- cophenetic(iris_es)
coph_ec <- cophenetic(iris_ec)
coph_ea <- cophenetic(iris_ea)
```

and show the cophenetic indices.

```{r cophenetic-indices}
(es <-cor(iris_e, coph_es))
(ec <-cor(iris_e, coph_ec))
(ea <-cor(iris_e, coph_ea))
```


## Extracting clusters

Try to extract different clusters!

```{r extract-clusters, exercise = TRUE}
k <- 2
(cluster_ea <- cutree(tree = iris_ea, k = k))
pairs(df, col = brewer.pal(n = ifelse(k >= 3, k, 3), name = "Accent")[cluster_ea])
```
