---
title: "Clustering - Earthquakes"
author: "Andrea Mascaretti"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE, exercise = FALSE)
```


# Earthquake Dataset


## Data Loading
We load the libraries and the data.
```{r load-libraries}
library(plotly)
```

```{r read-data}
df <- read.table(file = "../data/earthquakes.txt", header = TRUE)
```


## Plotting the data
```{r plot-data}
p <- plot_ly(df, x = ~lat, y = ~long, z = ~depth) %>%
  add_markers() %>%
  layout(scene = list(xaxis = list(title = 'Latitude'),
                     yaxis = list(title = 'Longitude'),
                     zaxis = list(title = 'Depth')))
p
```


## Rescaling the data
We rescale the data.
```{r rescale-fun}
rescale <- function(x, na.rm = FALSE) {
  (x - min(x, na.rm = na.rm)) / (max(x, na.rm = na.rm) - min(x, na.rm = na.rm))
}
```

```{r rescale-df}
df_r <- as.data.frame(apply(X = df, MARGIN = 2, FUN = rescale, na.rm = FALSE))
```


and we plot it.
```{r plot-data-rescaled}
p <- plot_ly(df_r, x = ~lat, y = ~long, z = ~depth) %>%
  add_markers() %>%
  layout(scene = list(xaxis = list(title = 'Latitude'),
                     yaxis = list(title = 'Longitude'),
                     zaxis = list(title = 'Depth')))
p
```



## k-means: finding k
```{r select-k}
within_ss <- matrix(data = NA, nrow = 1000, ncol = 10)
between_ss <- matrix(data = NA, nrow = 1000, ncol = 10)

for (k in seq.int(from = 1, to = 10)) {
  for (iter in seq.int(from = 1, to = 1000)) {
    clust_iter <- kmeans(x = df_r, centers = k)
    within_ss[iter, k] <- clust_iter$tot.withinss
    between_ss[iter, k] <- clust_iter$betweenss
  }
}
```

Let us check the result.
```{r select-k-within-plot}
boxplot(within_ss, main = "Within SS")
```
```{r select-k-between-plot}
boxplot(between_ss, main = "Between SS")
```


## k-means: clustering
```{r kmeans-clustering}
clust_k <- kmeans(x = df_r, centers = 4, iter.max = 50L, nstart = 100L)
```

We plot the results.
```{r kmeans-plot}
p <- plot_ly(df_r, x = ~lat, y = ~long, z = ~depth, color = ~as.factor(clust_k$cluster)) %>%
  add_markers() %>%
  layout(scene = list(xaxis = list(title = 'Latitude'),
                     yaxis = list(title = 'Longitude'),
                     zaxis = list(title = 'Depth')))
p
```


We also inspect what we have obtained.
```{r kmeans-inspection}
clust_k$iter
clust_k$cluster
clust_k$size
clust_k$centers
clust_k$tot.withinss
clust_k$betweenss
clust_k$totss
```



## Agglomerative Hierarchical Clustering
We now cluster considering with a different technique.

First of all, we compute the distance matrix.
```{r distance-matrix}
d <- dist(df_r, method = "euclidean")
```

```{r plot-heatmap}
image(as.matrix(d), asp = 1)
```


We perform different clusterings and plot them.
```{r hierarhical-compute}
clustc <- hclust(d, method='complete')
clusts <- hclust(d, method='single')
clusta <- hclust(d, method='average')
clustw <- hclust(d, method='ward.D')
```

```{r hierarchical-plot}
par(mfrow=c(2,2))
plot(clustc, hang=0, labels=FALSE, main='Complete', xlab='', sub='')
plot(clusts, hang=0, labels=FALSE, main='Single', xlab='', sub='')
plot(clusta, hang=0, labels=FALSE, main='Average', xlab='', sub='')
plot(clustw, hang=0, labels=FALSE, main='Ward', xlab='', sub='')
```


For each method, we shall display the results.


### Complete
```{r complete}
clusterc <- cutree(hclust(d, method='complete'), 2)
p <- plot_ly(df_r, x = ~lat, y = ~long, z = ~depth, color = ~as.factor(clusterc)) %>%
  add_markers() %>%
  layout(scene = list(xaxis = list(title = 'Latitude'),
                     yaxis = list(title = 'Longitude'),
                     zaxis = list(title = 'Depth')))
p
```
### Single
```{r single}
clusters <- cutree(hclust(d, method='single'), 4)
p <- plot_ly(df_r, x = ~lat, y = ~long, z = ~depth, color = ~as.factor(clusters)) %>%
  add_markers() %>%
  layout(scene = list(xaxis = list(title = 'Latitude'),
                     yaxis = list(title = 'Longitude'),
                     zaxis = list(title = 'Depth')))
p
```
### Average
```{r average}
clustera <- cutree(hclust(d, method='average'), 2)
p <- plot_ly(df_r, x = ~lat, y = ~long, z = ~depth, color = ~as.factor(clustera)) %>%
  add_markers() %>%
  layout(scene = list(xaxis = list(title = 'Latitude'),
                     yaxis = list(title = 'Longitude'),
                     zaxis = list(title = 'Depth')))
p
```
### Ward
```{r ward}
clusterw <- cutree(hclust(d, method='ward.D'), 3)
p <- plot_ly(df_r, x = ~lat, y = ~long, z = ~depth, color = ~as.factor(clusterw)) %>%
  add_markers() %>%
  layout(scene = list(xaxis = list(title = 'Latitude'),
                     yaxis = list(title = 'Longitude'),
                     zaxis = list(title = 'Depth')))
p
```